An efficient composition of bidirectional programs by tupling and lazy updates

### 2020-02-21 paper/1.introduction.tex
small changes: \circ -> \tilde{\circ} in intro. sec., \smallvspace

\begin{figure}[!t]
  \begin{minipage}{0.3\textwidth}
    \centering
    \includegraphics[height=3.5cm]{./fig/fig1.eps}
    \caption{Evaluating $phead$}
    \label{fig:eval-phead}
  \end{minipage}\hfill
  \begin{minipage}{0.7\textwidth}
    \centering
    \includegraphics[height=3.5cm]{./fig/fig2.eps}
    \caption{Evaluating $phead \circ phead \circ phead$}
    \label{fig:eval-comp-phead}
  \end{minipage}
\end{figure}
%\vspace*{-5.5mm}

The composition of BX programs is a fundamental construct to build more complex BX programs \cite{Bohannon06relationallenses:, Bohannon:2008:BRL:1328438.1328487}. Let $bx_1$ (defined by $get_{bx_1}$ and $put_{bx_1}$) and $bx_2$ (defined by $get_{bx_2}$ and $put_{bx_2}$) be two bidirectional programs, then their composition $bx_1 \circ bx_2$ is defined by

\minusvspace
\begin{align}
get_{bx_1 \circ bx_2}~s &= get_{bx_2} (get_{bx_1}~s)\\
put_{bx_1 \circ bx_2}~s~v' &= put_{bx_1} s~(put_{bx_2}~ (get_{bx_1}~s)~v')
\end{align}
\minusvspace

\noindent
\textcolor{red}{Unlike function composition, the composition of bidirectional programs is read left-to-right. We use this order because it is helpful to understand the behavior if we consider data goes from left to right.}
%\textcolor{red}{Note that the order of composition is the reverse order compared to function composition. We use this order, because it is standard in BX community.}
One feature of this composition is that $put_{bx_1 \circ bx_2}$ needs to call $get_{bx_1}$ to compute the intermediate result for $put_{bx_2}$ to use, which would introduce an efficiency problem if we compute $put$ for composition of many bidirectional programs. Generally, for a composition of $O(n)$ bidirectional programs, we need to call $get$ for $O(n^2)$ times. To be concrete, consider the evaluation of the following composition (which will be used as our running example in this paper):

\minusvspacetwo
\[
lp3 = (phead \circ phead) \circ phead
\]
\minusvspacetwo

@@

%In BX programs, composition of programs is essential for writing various interesting evaluations.
%Because recursions are achieved by compositions, large number of programs include compositions.
%In the BX lens papers \cite{Bohannon06relationallenses:, Bohannon:2008:BRL:1328438.1328487} compositions are also known as central part of BX programs.
%In this section we use a small BX program that contains compositions as a running example, because the programs including recursions can be complicated easily. The running examples are two versions of two composition of $phead$: ($phead \circ phead) \circ phead$ (we call lp3) and $phead \circ (phead \circ phead$) (we call rp3).

%For more various evaluations, we can use the composition of programs. Let us consider another example, a composition of 3 $pHead$s: $pHead \circ pHead \circ pHead$. Figure \ref{fig:eval-comp-phead} illustrates the evaluation of this program where the original source \texttt{s} is \texttt{[[[1,2],[3]],[4]]} and the updated view is \texttt{100}.
%By repetitious evaluation of $get$s of $pHead$, we can obtain the view $v$, \texttt{1}. To obtain the final updated source \texttt{s'}, we need to evaluate $put$ three times. The first put evaluation is from \texttt{i2} and \texttt{v'} and we can obtain \texttt{i2'}.

@@

\minusvspacetwo
\vspace{-3mm}
\[
rp3 = phead \circ (phead \circ phead)
\]
\minusvspacetwo
\vspace{-0mm}

@@

\vspace{1mm}
\minusvspacetwo
 \[
 \mbox{\it bfoldr}~bx~\cdots = \cdots  (\mbox{\it bfoldr}~\cdots) \circ bx \cdots
 \]
 \minusvspacetwo

@@

%A part of the definition of $\mbox{\it bfoldr}$, we have $(\product{Replace}{(\mbox{\it bfoldr} \ bx)}) \circ bx$. In this program, recursion occurs on the left-hand side of composition. Because of priority, it is impossible to transform this program to right associative compositions. At the same time, we can not apply program fusion \cite{Wadler:1988:DTP:80099.80104} to this program because actual program is dynamically produced by recursions depending on the length of the input list. Therefore, it is important to have a fast evaluation method for left associative compositions.

\begin{figure}[!t]
  \centering
  \includegraphics[height=5cm]{./fig/fig3.eps}
  \vspace{-0.25cm}
  \caption{Evaluating $phead \ \circ \ phead \ \circ \ phead$ by keeping complements}
  \vspace{-0.5cm}
  \label{fig:eval-comp-phead-2}
\end{figure}

@@

To treat large inputs, we propose the second method based on memoization: ``keeping \emph{complements in a closure} and using them when needed''.
\textcolor{red}{Here, complements are information from sources that makes $get$ injective, which is in turn needed to evaluate $put$.
  In the middle $put$ and $get$ of Figure \ref{fig:eval-comp-phead},
  we use $i_1$, $[[1,2],[3]]$ and $i_2'$, $[100,2]$ to obtain the updated source $[[100,2],[3]]$. However, $[1,2]$ is simply replaced by $[100,2]$ and not used to construct the result of $put$. In this case we can use $[..., [3]]$ as a complement.}
The key idea of the second approach is straightforward: Complements are smaller
% program fragments
than
% the original
intermediate states.
%Therefore this strategy solves the previous two problems.
%Readers might already know that it is a hard problem to obtain complements in general. In our work, thanks to the following finding, it is not hard:
%
%\vspace{2mm}
%In very-well behaved \cite{Foster:2007:CBT:1232420.1232424} BX,% programs,
%$put$ is a complement function for $get$.
%, $get$ can be a complement function for $put$.
%\vspace{2mm}
%
For obtaining complements, we tuple $put$ and $get$, and produce a new function $pg$. Because $put$ produces new complements for $get$, we can shrink the size.
Let us reconsider our example $phead \circ phead \circ phead$ in Figure \ref{fig:eval-comp-phead-2}, where $c_1$ and $c_2$ are complements and $d_1$ and $d_2$ are valid views for $s$ and~$i_1$. Here, two points are worth noting. First, after evaluating the leftmost $pg$, the original source $s$ need not be kept because its complete contents are in $c_1$ and~$i_1$. Second, the complements are smaller than the intermediate states in Figure~\ref{fig:eval-comp-phead}.
%This evaluation looks better than previous two strategies: this does not require repeated evaluation and require smaller storage than the original sources.
Actually, this simple
% combined
$pg$ alone is not yet effective for left associative compositions because it requires two more $put$s, which can be seen on the right side of Figure \ref{fig:eval-comp-phead-2}. To achieve an efficient evaluation, we combine two techniques, lazy update and lazy evaluation. We explain this second method and all optimizations in Section~\ref{sect:xpg}.

>>>>

\begin{figure}[!t]
  \begin{minipage}{0.3\textwidth}
    \centering
    \includegraphics[height=3.5cm]{./fig/fig1.eps}
    \caption{Evaluating $phead$}
    \label{fig:eval-phead}
  \end{minipage}\hfill
  \begin{minipage}{0.7\textwidth}
    \centering
    \includegraphics[height=3.5cm]{./fig/fig2.eps}
    \caption{Evaluating $phead~\tilde{\circ}~phead~\tilde{\circ}~phead$}
    \label{fig:eval-comp-phead}
  \end{minipage}
\end{figure}
%\vspace*{-5.5mm}

The composition of BX programs is a fundamental construct to build more complex BX programs \cite{Bohannon06relationallenses:, Bohannon:2008:BRL:1328438.1328487}. Let $bx_1$ (defined by $get_{bx_1}$ and $put_{bx_1}$) and $bx_2$ (defined by $get_{bx_2}$ and $put_{bx_2}$) be two bidirectional programs, then their composition $bx_1 \,\tilde{\circ}\, bx_2$ is defined by

\minusvspace
\begin{align}
get_{bx_1 \tilde{\circ} bx_2}~s &= get_{bx_2} (get_{bx_1}~s)\\
put_{bx_1 \tilde{\circ} bx_2}~s~v' &= put_{bx_1} s~(put_{bx_2}~ (get_{bx_1}~s)~v')
\end{align}
\minusvspace

\noindent
\textcolor{red}{Unlike function composition, the composition of bidirectional programs is read left-to-right. We use this order because it is helpful to understand the behavior if we consider data goes from left to right.}
%\textcolor{red}{Note that the order of composition is the reverse order compared to function composition. We use this order, because it is standard in BX community.}
One feature of this composition is that $put_{bx_1 \tilde{\circ} bx_2}$ needs to call $get_{bx_1}$ to compute the intermediate result for $put_{bx_2}$ to use, which would introduce an efficiency problem if we compute $put$ for composition of many bidirectional programs. Generally, for a composition of $O(n)$ bidirectional programs, we need to call $get$ for $O(n^2)$ times. To be concrete, consider the evaluation of the following composition (which will be used as our running example in this paper):

\minusvspacetwo
\[
lp3 = (phead~\tilde{\circ}~phead)~\tilde{\circ}~phead
\]
\minusvspacetwo

@@

%In BX programs, composition of programs is essential for writing various interesting evaluations.
%Because recursions are achieved by compositions, large number of programs include compositions.
%In the BX lens papers \cite{Bohannon06relationallenses:, Bohannon:2008:BRL:1328438.1328487} compositions are also known as central part of BX programs.
%In this section we use a small BX program that contains compositions as a running example, because the programs including recursions can be complicated easily. The running examples are two versions of two composition of $phead$: ($phead \tilde{\circ} phead) \tilde{\circ} phead$ (we call lp3) and $phead \tilde{\circ} (phead \tilde{\circ} phead$) (we call rp3).

%For more various evaluations, we can use the composition of programs. Let us consider another example, a composition of 3 $pHead$s: $pHead \tilde{\circ} pHead \tilde{\circ} pHead$. Figure \ref{fig:eval-comp-phead} illustrates the evaluation of this program where the original source \texttt{s} is \texttt{[[[1,2],[3]],[4]]} and the updated view is \texttt{100}.
%By repetitious evaluation of $get$s of $pHead$, we can obtain the view $v$, \texttt{1}. To obtain the final updated source \texttt{s'}, we need to evaluate $put$ three times. The first put evaluation is from \texttt{i2} and \texttt{v'} and we can obtain \texttt{i2'}.

@@

\minusvspacetwo
\vspace{-3mm}
\[
rp3 = phead~\tilde{\circ}~(phead~\tilde{\circ}~phead)
\]
\minusvspacetwo
\vspace{-0mm}

@@

\vspace{1mm}
\minusvspacetwo
 \[
 \mbox{\it bfoldr}~bx~\cdots = \cdots  (\mbox{\it bfoldr}~\cdots)~\tilde{\circ}~bx \cdots
 \]
 \minusvspacetwo

@@

%A part of the definition of $\mbox{\it bfoldr}$, we have $(\product{Replace}{(\mbox{\it bfoldr} \ bx)}) \tilde{\circ} bx$. In this program, recursion occurs on the left-hand side of composition. Because of priority, it is impossible to transform this program to right associative compositions. At the same time, we can not apply program fusion \cite{Wadler:1988:DTP:80099.80104} to this program because actual program is dynamically produced by recursions depending on the length of the input list. Therefore, it is important to have a fast evaluation method for left associative compositions.

\begin{figure}[!t]
  \centering
  \includegraphics[height=5cm]{./fig/fig3.eps}
  \vspace{-0.25cm}
  \caption{Evaluating $phead \ \tilde{\circ} \ phead \ \tilde{\circ} \ phead$ by keeping complements}
  \vspace{-0.5cm}
  \label{fig:eval-comp-phead-2}
\end{figure}

@@

To treat large inputs, we propose the second method based on memoization: ``keeping \emph{complements in a closure} and using them when needed''.
\textcolor{red}{Here, complements are information from sources that makes $get$ injective, which is in turn needed to evaluate $put$.
  In the middle $put$ and $get$ of Figure \ref{fig:eval-comp-phead},
  we use $i_1$, $[[1,2],[3]]$ and $i_2'$, $[100,2]$ to obtain the updated source $[[100,2],[3]]$. However, $[1,2]$ is simply replaced by $[100,2]$ and not used to construct the result of $put$. In this case we can use $[..., [3]]$ as a complement.}
The key idea of the second approach is straightforward: Complements are smaller
% program fragments
than
% the original
intermediate states.
%Therefore this strategy solves the previous two problems.
%Readers might already know that it is a hard problem to obtain complements in general. In our work, thanks to the following finding, it is not hard:
%
%\vspace{2mm}
%In very-well behaved \cite{Foster:2007:CBT:1232420.1232424} BX,% programs,
%$put$ is a complement function for $get$.
%, $get$ can be a complement function for $put$.
%\vspace{2mm}
%
For obtaining complements, we tuple $put$ and $get$, and produce a new function $pg$. Because $put$ produces new complements for $get$, we can shrink the size.
Let us reconsider our example $phead~\tilde{\circ}~phead~\tilde{\circ}~phead$ in Figure \ref{fig:eval-comp-phead-2}, where $c_1$ and $c_2$ are complements and $d_1$ and $d_2$ are valid views for $s$ and~$i_1$. Here, two points are worth noting. First, after evaluating the leftmost $pg$, the original source $s$ need not be kept because its complete contents are in $c_1$ and~$i_1$. Second, the complements are smaller than the intermediate states in Figure~\ref{fig:eval-comp-phead}.
%This evaluation looks better than previous two strategies: this does not require repeated evaluation and require smaller storage than the original sources.
Actually, this simple
% combined
$pg$ alone is not yet effective for left associative compositions because it requires two more $put$s, which can be seen on the right side of Figure \ref{fig:eval-comp-phead-2}. To achieve an efficient evaluation, we combine two techniques, lazy update and lazy evaluation. We explain this second method and all optimizations in Section~\ref{sect:xpg}.

### 2020-02-21 paper/2.minbigul.tex
small changes: \circ -> \tilde{\circ} in intro. sec., \smallvspace

To help make demonstration more direct, we provide the following alternatives representation: $Prod \ bx_1 \ bx_2 \equiv bx_1 \times bx_2, \ Compose \ bx_1 \ bx_2 \equiv bx_1 \circ bx_2$. In general, $\circ$ has a higher priority than $\times$. Their associativity precedence can be either left or right or mixture, but are not set by default. We need to explicitly write programs that use these operators.

@@

\noindent As an example of minBiGUL program, consider the definition of $\textit{phead}$ :\\
\smallvspace
$\tab phead = RearrS \ f_1 \ f_2 \ bx_s \ \text{where: } f_1 = \lambda (s::ss).(s,ss), \, f_2 = \lambda (s,ss).(s::ss),\\
    \qtab bx_s = RearrV \ g_1 \ g_2 \ bx_v \ \text{where: } g_1 = \lambda v.(v,()), \, g_2 = \lambda (v,()).v,\\
        \qtabs{2} bx_v = \product{Replace}{(Skip \ (\lambda \_.())}$
\smallvspace

@@

    $\tab \textit{bsnoc} = Case \ cond_{sv} \ cond_s \ bx_1 \ bx_2$ where:\\
    $\tabs{3} \ cond_{sv} = \lambda s.\lambda v.(\textnormal{length } v = 1), \ cond_s = \lambda s.(\textnormal{length } s = 1)$\\
    $\tabs{3} \ bx_1 =  Replace, \tab bx_2 = RearrS \ f_1 \ f_2 \ bx_s$ where:\\
        $\tabs{4} f_1 = f_2^{-1} = \lambda (x:y:ys).(y,(x:ys))$, \\
        $\tabs{4} bx_s = RearrV \ g_1 \ g_2 \ bx_v$ where:\\
            $\tabs{5} g_1 = g_2^{-1} = \lambda (v:vs).(v,vs), \tab bx_v = Replace \times \textit{bsnoc}$

@@

$\tab \textit{bfoldr} \ bx = Case \ cond_{sv} \ cond_s \ bx_1 \ bx_2$ where:\\
$\tabs{3} \ cond_{sv} = \lambda (s_1,s_2).\lambda v.(s_1 = [\ ]), \ cond_s = \lambda (s_1,s_2).(s_1 = [\ ])$\\
$\tabs{3} \ bx_1 = RearrV \ g_1 \ g_2 \ bx_v$ where: \\
    $\tabs{4} g_1 = g_2^{-1} = \lambda [v].(v,[\ ]), bx_v = (Skip \ (\lambda \_.())) \times Replace$\\
$\tabs{3} \ bx_2 = RearrS \ f_1 \ f_2 \ bx_s$ where:\\
    $\tabs{4} f_1 = f_2^{-1} = \lambda ((x:xs),e).(x,(xs,e))$, 
$bx_s = ((Replace \times \textit{bfoldr} \ bx) \circ bx)$

@@

$\tab \textit{foldr} :: (a \to b \to b) \to b \to [a] \to b \\
\tabs{2} \ \ \textit{foldr} \ f \ e \ [\ ] = e \\ 
\tabs{2} \ \ \textit{foldr} \ f \ e \ (x:xs) = f \ x \ (\textit{foldr} \ f \ e \ xs)$

@@

$\tab breverse = RearrS \ f_1 \ f_2 \ bx$ where: \\
$\tabs{4} f_1 = f_2^{-1} = \lambda s \to (s, [\ ]), \tab bx = \textit{bfoldr} \ bsnoc$

>>>>

To help make demonstration more direct, we provide the following alternatives representation: $Prod \ bx_1 \ bx_2 \equiv bx_1 \times bx_2, \ Compose \ bx_1 \ bx_2 \equiv bx_1 \circ bx_2$. \textcolor{blue}{The compose symbol $\tilde{\circ}$ used in the previous section will be replaced with the more common one, $\circ$.}
In general, $\circ$ has a higher priority than $\times$. Their associativity precedence can be either left or right or mixture, but are not set by default. We need to explicitly write programs that use these operators.

@@

\noindent As an example of minBiGUL program, consider the definition of $\textit{phead}$:

\smallvspace
\noindent $\tab phead = RearrS \ f_1 \ f_2 \ bx_s \ \text{where: } f_1 = \lambda (s::ss).(s,ss), \, f_2 = \lambda (s,ss).(s::ss),\\
    \qtab bx_s = RearrV \ g_1 \ g_2 \ bx_v \ \text{where: } g_1 = \lambda v.(v,()), \, g_2 = \lambda (v,()).v,\\
        \qtabs{2} bx_v = \product{Replace}{(Skip \ (\lambda \_.())}$
\smallvspace

@@

\smallvspace
    $\tab \textit{bsnoc} = Case \ cond_{sv} \ cond_s \ bx_1 \ bx_2$ where:\\
    $\tabs{3} \ cond_{sv} = \lambda s.\lambda v.(\textnormal{length } v = 1), \ cond_s = \lambda s.(\textnormal{length } s = 1)$\\
    $\tabs{3} \ bx_1 =  Replace, \tab bx_2 = RearrS \ f_1 \ f_2 \ bx_s$ where:\\
        $\tabs{4} f_1 = f_2^{-1} = \lambda (x:y:ys).(y,(x:ys))$, \\
        $\tabs{4} bx_s = RearrV \ g_1 \ g_2 \ bx_v$ where:\\
            $\tabs{5} g_1 = g_2^{-1} = \lambda (v:vs).(v,vs), \tab bx_v = Replace \times \textit{bsnoc}$
\smallvspace

@@

\smallvspace
$\tab \textit{bfoldr} \ bx = Case \ cond_{sv} \ cond_s \ bx_1 \ bx_2$ where:\\
$\tabs{3} \ cond_{sv} = \lambda (s_1,s_2).\lambda v.(s_1 = [\ ]), \ cond_s = \lambda (s_1,s_2).(s_1 = [\ ])$\\
$\tabs{3} \ bx_1 = RearrV \ g_1 \ g_2 \ bx_v$ where: \\
    $\tabs{4} g_1 = g_2^{-1} = \lambda [v].(v,[\ ]), bx_v = (Skip \ (\lambda \_.())) \times Replace$\\
$\tabs{3} \ bx_2 = RearrS \ f_1 \ f_2 \ bx_s$ where:\\
    $\tabs{4} f_1 = f_2^{-1} = \lambda ((x:xs),e).(x,(xs,e))$, 
$bx_s = ((Replace \times \textit{bfoldr} \ bx) \circ bx)$
\smallvspace

@@

\smallvspace
$\tab \textit{foldr} :: (a \to b \to b) \to b \to [a] \to b \\
\tabs{2} \ \ \textit{foldr} \ f \ e \ [\ ] = e \\ 
\tabs{2} \ \ \textit{foldr} \ f \ e \ (x:xs) = f \ x \ (\textit{foldr} \ f \ e \ xs)$
\smallvspace

@@

\smallvspace
$\tab breverse = RearrS \ f_1 \ f_2 \ bx$ where: \\
$\tabs{4} f_1 = f_2^{-1} = \lambda s \to (s, [\ ]), \tab bx = \textit{bfoldr} \ bsnoc$
